### 2023-09-03 sections/1-introduction.tex
misc changes

Biology has also seen an influx of large, high-dimensional datasets. Silva 18S~\cite{10.1093/nar/gks1219} contains ribosomal DNA sequences of approximately 2.25 
million genomes with aligned length of 50,000 letters; 
the GreenGenes project~\cite{desantis2006greengenes} provides a multiple-sequence alignment of over one million bacterial 16S sequences. [tTODO: add dimensionality here]

@@

Many researchers are especially interested in similarity search on these datasets. 
Similarity search enables a variety of applications, including recommendation and classification systems. 
Additionally, with the rise of large language models such as GPT-3~\cite{2020arXiv200514165B}, GPT-4, and LLAMA-2, [CITE] similarity search can become a useful tool for 
understanding and verifying the validity of model embeddings. 
As the sizes and dimensionalities of datasets have grown, however, efficient and accurate similarity search has become extremely challenging; 
even state-of-the-art algorithms exhibit a steep recall versus throughput tradeoff~\cite{ishaq2019clustered}.

>>>>

Biology has also seen an influx of large, high-dimensional datasets. Silva 18S~\cite{10.1093/nar/gks1219} contains ribosomal DNA sequences of approximately 2.25 
million genomes with aligned length of 50,000 letters; 
the GreenGenes project~\cite{desantis2006greengenes} provides a multiple-sequence alignment of over one million bacterial 16S sequences, each 7,682 characters in length.

@@

Many researchers are especially interested in similarity search on these datasets. 
Similarity search enables a variety of applications, including recommendation and classification systems. 
Additionally, with the rise of large language models such as GPT-3~\cite{2020arXiv200514165B}, GPT-4~\cite{OpenAI2023GPT4TR}, and LLAMA-2~\cite{Touvron2023Llama2O}, similarity search can become a useful tool for 
understanding and verifying the validity of model embeddings. 
As the sizes and dimensionalities of datasets have grown, however, efficient and accurate similarity search has become extremely challenging; 
even state-of-the-art algorithms exhibit a steep recall versus throughput tradeoff~\cite{ishaq2019clustered}.

### 2023-09-03 sections/2-methods.tex
misc changes

[TODO: change to just partition because I took about quicksort partition]
\subsubsection {Cluster Partitioning}

@@

\begin{algorithm} % enter the algorithm environment
\caption{ClusterPartition(\emph{C})} % give the algorithm a caption
\label{alg:partition} % and a label for \ref{} commands later in the document
\begin{algorithmic} % enter the algorithmic environment
    \STATE $m \Leftarrow \lfloor \sqrt{|C|} \rfloor$
    \STATE $seeds \Leftarrow m$ random points from $C.points$
    \STATE $c \Leftarrow$ geometric median of $seeds$
    \STATE $l \Leftarrow \argmax d(c,x) \ \forall \ x \in C.points$
    \STATE $r \Leftarrow \argmax d(l,x) \ \forall \ x \in C.points$
    \STATE $left \Leftarrow \{x | x \in C.points \land d(l,x) \le d(r,x)\}$
    \STATE $right \Leftarrow \{x | x \in C.points \land d(r,x) < d(l,x)\}$
    \IF{$|left| > 1$}
        \STATE ClusterPartition($left$)
    \ENDIF
    \IF{$|right| > 1$}
        \STATE ClusterPartition($right$)
    \ENDIF
\end{algorithmic}
\end{algorithm}

@@

Note that, in practice, we never see a balanced clustering. Cluster Partition  produces unbalanced trees due to the varying density of the sampling 
in different regions of the manifold and the low dimensional ``shape'' of the manifold. Further, in practice, the cluster radii decrease by a factor much larger than 
$\frac{\sqrt{2}}{2}$ with every partition; the upper bound of $d$ partitions is seldom realized.

>>>>

\subsubsection {Cluster Partitioning}

@@

\begin{algorithm} % enter the algorithm environment
\caption{Partition(\emph{C})} % give the algorithm a caption
\label{alg:partition} % and a label for \ref{} commands later in the document
\begin{algorithmic} % enter the algorithmic environment
    \STATE $m \Leftarrow \lfloor \sqrt{|C|} \rfloor$
    \STATE $seeds \Leftarrow m$ random points from $C.points$
    \STATE $c \Leftarrow$ geometric median of $seeds$
    \STATE $l \Leftarrow \argmax d(c,x) \ \forall \ x \in C.points$
    \STATE $r \Leftarrow \argmax d(l,x) \ \forall \ x \in C.points$
    \STATE $left \Leftarrow \{x | x \in C.points \land d(l,x) \le d(r,x)\}$
    \STATE $right \Leftarrow \{x | x \in C.points \land d(r,x) < d(l,x)\}$
    \IF{$|left| > 1$}
        \STATE Partition($left$)
    \ENDIF
    \IF{$|right| > 1$}
        \STATE Partition($right$)
    \ENDIF
\end{algorithmic}
\end{algorithm}

@@

Note that, in practice, we never see a balanced clustering. Partition  produces unbalanced trees due to the varying density of the sampling 
in different regions of the manifold and the low dimensional ``shape'' of the manifold. Further, in practice, the cluster radii decrease by a factor much larger than 
$\frac{\sqrt{2}}{2}$ with every partition; the upper bound of $d$ partitions is seldom realized.
